---
title: "The fqar package"
author: "Andrew Gard and Alexia Myers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fqar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, results = "hide"}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fqar)
library(ggplot2)
library(dplyr)
```

## Introduction

The ${\tt fqar}$ packages provides tools for downloading and analyzing floristic quality assessment (FQA) data from [universalFQA.org](https://universalfqa.org/). Two sample data sets, ${\tt missouri}$ and ${\tt chicago}$, are also provided.

Functions in this package fall into three general categories: _indexing functions_, which produce data frames of current public databases and FQAs from various regions, _downloading functions_, which download the FQAs themselves, and _tidying functions_, which convert downloaded assessments into a standard format.

## Indexing functions

Each floristic quality assessment is tied to a specific databases of native plants that has been compiled by experts in local flora. A listing of all databases accepted by [universalFQA.org](https://universalfqa.org/) can be viewed with the ${\tt index\_fqa\_databases()}$ function.

```{r}
databases <- index_fqa_databases()
head(databases)
```

To see a listing of all public floristic quality assessments using a given database, use the ${\tt index\_fqa\_assessments()}$ function.

```{r}
missouri_fqas <- index_fqa_assessments(database_id = 63)
head(missouri_fqas)
```

Similarly, the ${\tt index\_fqa\_transects()}$ function returns a listing of all public transect assessments using the specified database.  

```{r}
missouri_transects <- index_fqa_transects(database_id = 63)
head(missouri_transects)
```

## Downloading functions

Floristic quality assessments can be downloaded individually by id number or in batches according to specified search criteria using the ${\tt download\_assessment()}$ and ${\tt download\_assessment\_list()}$ functions, respectively. 

The first of these accepts an assessment ID number as its sole input and returns a data frame. For instance, the Grasshopper Hollow survey has ${\tt assessment\_id} = 25961$ according to the listing obtained using ${\tt index\_fqa\_assessments()}$. The following code downloads this assessment.

```{r}
grasshopper <- download_assessment(assessment_id = 25961)
```

Multiple assessments from a specified database can be downloaded simultaneously using ${\tt download\_assessment\_list()}$, which makes use of ${\tt dplyr::filter}$ syntax on the variables ${\tt id, assessment, date, site}$ and ${\tt practitioner}$. For instance, the following code downloads all assessments performed using the 2015 Missouri database at the Ambrose Farm site.

```{r}
ambrose <- download_assessment_list(database_id = 63,
                                    site == "Ambrose Farm")
```

For even mid-sized requests, this command may run slowly due to the limited speed of the [universalFQA.org](https://universalfqa.org/) website. For this reason, a progress bar has been added to the ${\tt download\_assessment\_list()}$ function when $n\ge 5$. 

As the name suggests, the output of ${\tt download\_assessment\_list()}$ is a list of data frames. 

```{r}
class(ambrose)
length(ambrose)
```

Transect assessment data data stored on [universalFQA.org](https://universalfqa.org/) is accessible to analysts using the ${\tt fqar}$ package via the functions ${\tt download\_transect()}$ and ${\tt download\_transect\_list()}$, which work exactly like their counterparts, ${\tt download\_assessment()}$ and ${\tt download\_assessment\_list()}$.

```{r}
rock_garden <- download_transect(transect_id = 6875)
golden <- download_transect_list(database_id = 63,
                                 site == "Golden Prairie")
```

## Tidying functions

The data frames obtained from these downloading functions are all highly untidy, respecting the default structure of the website from which they are obtained. The ${\tt fqar}$ package provides tools for efficiently re-formatting these sets.

Each floristic quality assessments on [universalFQA.org](https://universalfqa.org/) includes two types of information: details about the species observed during data collection and summary information about the assessment as a whole. The ${\tt fqar}$ functions ${\tt assessment\_inventory()}$ and ${\tt assessment\_glance()}$ extract and tidy these two types of information.

For instance, the following code creates a data frame of species found in the 2021 Grasshopper Hollow survey downloaded earlier.

```{r}
grasshopper_species <- assessment_inventory(grasshopper)
glimpse(grasshopper_species)
```

A tidy summary of the assessment can be obtained with ${\tt assessment\_glance()}$. The output is a data frame with a single row and 53 columns, including ${\tt native\_mean\_C}$,  ${\tt native\_species}$, and ${\tt native\_fqi}$.

```{r}
grasshopper_summary <- assessment_glance(grasshopper)
names(grasshopper_summary)
```

The tidy format provided by ${\tt assessment\_glance()}$ is most useful when applied to multiple data sets at once, for instance in the situation where the analyst wants to consider statistics from many different assessments simultaneously. The ${\tt assessment\_list\_glance()}$ function provides a shortcut when those data frames are housed in a list like that returned by ${\tt download\_assessment\_list()}$. For instance, the following code returns a data frame with 52 columns and 3 rows, one per assessment.

```{r}
ambrose_summary <- assessment_list_glance(ambrose)
```

The ${\tt fqar}$ package also provides functions for handling transect assessment data. ${\tt transect\_inventory()}$, ${\tt transect\_glance()}$, and ${\tt transect\_list\_glance()}$ work just like their counterparts, ${\tt assessment\_inventory()}$, ${\tt assessment\_glance()}$, and ${\tt assessment\_list\_glance()}$.

```{r, eval = FALSE}
rock_garden_species <- transect_inventory(rock_garden)
rock_garden_summary <- transect_glance(rock_garden)
golden_summary <- transect_list_glance(golden)
```

Additionally, transect assessments usually include physiognometric metrics like relative frequency and relative coverage. These can be extracted with the ${\tt trasect\_phys()}$ function.

```{r}
rock_garden_phys <- transect_phys(rock_garden)
glimpse(rock_garden_phys)
```

## Analytic functions

The ${\tt fqar}$ package provides tools for analyzing species co-occurrence across multiple floristic quality assessments. A typical workflow consists of downloading a list of assessments, extracting inventories from each, then enumerating and summarizing co-occurrences of the species of interest.

```{r analysis}
# Obtain a tidy data frame of all co-occurrences in the 1995 Southern Ontario database:
ontario <- download_assessment_list(database = 2)

# Extract inventories as a list:
ontario_invs <- assessment_list_inventory(ontario)

# Enumerate all co-occurrences in this database:
ontario_cooccurrences <- assessment_cooccurrences(ontario_invs)

# Sumamrize co-occurrences in this database, one row per target species:
ontario_cooccurrences <- assessment_cooccurrences_summary(ontario_invs)
```

Of particular note is the ${\tt species\_profile()}$ function, which returns the frequency distribution of C-values of co-occurring species for a given target species. Users may specify the optional ${\tt native}$ argument to include only native species in the profile. 

For instance, Aster lateriflorus (C=0) has the following profile. 

```{r profile}
aster_profile <- species_profile("Aster lateriflorus", 
                                 ontario_invs,
                                 native = TRUE)

ggplot(aster_profile, aes(x = cospecies_c, y = cospecies_n)) +
  geom_col() +
  scale_x_continuous(breaks = seq(from = 0, to = 11, by = 2)) +
  labs(x = "Co-occurring species C values",
       y = "Frequency",
       title = paste(aster_profile$species, "native co-occurrence profile")) +
  geom_vline(xintercept = aster_profile$target_c,
             linetype = "dashed") +
  theme_minimal()
```

## Data sets

Two tidy data sets of floristic quality data, ${\tt missouri}$ and ${\tt chicago}$, are included with the ${\tt fqar}$ package. Produced with ${\tt assessment\_list\_glance()}$, these show summary information for every floristic quality assessment that used databases 63 and 149, respectively, prior to August 14, 2022. These sets may be useful for visualization or machine-learning purposes. For instance, one might consider the relationship between richness and native mean C in sites assessed using the 2015 Missouri database:

```{r missouri plot, warning = FALSE}
ggplot(missouri, aes(x = native_species, 
                     y = native_mean_c)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(trans = "log10") +
  labs(x = "Native Species (logarithmic scale)",
       y = "Native Mean C") +
  theme_minimal()
```

