---
title: "The fqar package"
author: "Andrew Gard and Alexia Myers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fqar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, results = "hide"}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fqar)
library(ggplot2)
library(dplyr)
```

## Introduction

The ${\tt fqar}$ packages provides tools for downloading and analyzing floristic quality assessment (FQA) data from [universalFQA.org](https://universalfqa.org/). Two sample data sets, `chicago` and  `missouri`, are also provided.

Functions in this package fall into four general categories: _indexing functions_, which produce data frames of current public databases and FQAs from various regions, _downloading functions_, which download the FQAs themselves, _tidying functions_, which convert downloaded assessments into a standard format, and _analytic functions_, which compare species across assessments.

## Indexing functions

Each floristic quality assessment is tied to a specific databases of native plants that has been compiled by experts in local flora. A listing of all databases accepted by [universalFQA.org](https://universalfqa.org/) can be viewed with the `index_fqa_databases()` function.

```{r}
databases <- index_fqa_databases()
head(databases)
```

To see a listing of all public floristic quality assessments using a given database, use the `index_fqa_assessments()` function.

```{r}
missouri_fqas <- index_fqa_assessments(database_id = 63)
head(missouri_fqas)
```

Similarly, the `index_fqa_transects()` function returns a listing of all public transect assessments using the specified database.  

```{r}
missouri_transects <- index_fqa_transects(database_id = 63)
head(missouri_transects)
```

## Downloading functions

Floristic quality assessments can be downloaded individually by id number or in batches according to specified search criteria using the `download_assessment()` and `download_assessment_list()` functions, respectively. 

The first of these accepts an assessment ID number as its sole input and returns a data frame. For instance, the Grasshopper Hollow survey has `assessment_id = 25961` according to the listing obtained using `index_fqa_assessments()`. The following code downloads this assessment.

```{r}
grasshopper <- download_assessment(assessment_id = 25961)
```

Multiple assessments from a specified database can be downloaded simultaneously using `download_assessment_list()`, which makes use of `dplyr::filter` syntax on the variables `id, assessment, date, site` and `practitioner`. For instance, the following code downloads all assessments performed using the 2015 Missouri database at the Ambrose Farm site.

```{r}
ambrose <- download_assessment_list(database_id = 63,
                                    site == "Ambrose Farm")
```

For even mid-sized requests, this command may run slowly due to the limited speed of the [universalFQA.org](https://universalfqa.org/) website. For this reason, a progress bar has been added to the `download_assessment_list()` function when $n\ge 5$. 

As the name suggests, the output of `download_assessment_list()` is a list of data frames. 

```{r}
class(ambrose)
length(ambrose)
```

Transect assessment data data stored on [universalFQA.org](https://universalfqa.org/) is accessible to analysts using the ${\tt fqar}$ package via the functions `download_transect()` and `download_transect_list()`, which work exactly like their counterparts, `download_assessment()` and `download_assessment_list()`.

```{r}
rock_garden <- download_transect(transect_id = 6875)
golden <- download_transect_list(database_id = 63,
                                 site == "Golden Prairie")
```

## Tidying functions

The data frames obtained from these downloading functions are all highly untidy, respecting the default structure of the website from which they are obtained. The ${\tt fqar}$ package provides tools for efficiently re-formatting these sets.

Each floristic quality assessments on [universalFQA.org](https://universalfqa.org/) includes two types of information: details about the species observed during data collection and summary information about the assessment as a whole. The ${\tt fqar}$ functions `assessment_inventory()` and `assessment_glance()` extract and tidy these two types of information.

For instance, the following code creates a data frame of species found in the 2021 Grasshopper Hollow survey downloaded earlier.

```{r}
grasshopper_species <- assessment_inventory(grasshopper)
glimpse(grasshopper_species)
```

A tidy summary of the assessment can be obtained with `assessment_glance()`. The output is a data frame with a single row and 53 columns, including `native_mean_c`,  `native_species`, and `native_fqi`.

```{r}
grasshopper_summary <- assessment_glance(grasshopper)
names(grasshopper_summary)
```

The tidy format provided by `assessment_glance()` is most useful when applied to multiple data sets at once, for instance in the situation where the analyst wants to consider statistics from many different assessments simultaneously. The `assessment_list_glance()` function provides a shortcut when those data frames are housed in a list like that returned by `download_assessment_list()`. For instance, the following code returns a data frame with 52 columns and 3 rows, one per assessment.

```{r}
ambrose_summary <- assessment_list_glance(ambrose)
```

The ${\tt fqar}$ package also provides functions for handling transect assessment data. `transect_inventory()`, transect_glance()`, and `transect_list_glance()` work just like their counterparts, `assessment_inventory()`, `assessment_glance()`, and `assessment_list_glance()`.

```{r, eval = FALSE}
rock_garden_species <- transect_inventory(rock_garden)
rock_garden_summary <- transect_glance(rock_garden)
golden_summary <- transect_list_glance(golden)
```

Additionally, transect assessments usually include physiognometric metrics like relative frequency and relative coverage. These can be extracted with the `trasect_phys()` function.

```{r}
rock_garden_phys <- transect_phys(rock_garden)
glimpse(rock_garden_phys)
```

## Analytic functions

The ${\tt fqar}$ package provides tools for analyzing species co-occurrence across multiple floristic quality assessments. A typical workflow consists of downloading a list of assessments, extracting inventories from each, then enumerating and summarizing co-occurrences of species of interest.

```{r analysis, results = 'hide'}
# Obtain a tidy data frame of all co-occurrences in the 1995 Southern Ontario database:
ontario <- download_assessment_list(database = 2)

# Extract inventories as a list:
ontario_invs <- assessment_list_inventory(ontario)

# Enumerate all co-occurrences in this database:
ontario_cooccurrences <- assessment_cooccurrences(ontario_invs)

# Summarize co-occurrences in this database, one row per target species:
ontario_cooccurrences <- assessment_cooccurrences_summary(ontario_invs)
```

Of particular note is the `species_profile()` function, which returns the frequency distribution of C-values of co-occurring species for a given target species. Users may specify the optional `native` argument to include only native species in the profile. The `species_profile_plot()` function takes identical arguments but returns an elegant plot instead of a data frame

For instance, _Aster lateriflorus_ (C=3) has the following native profile in the Southern Ontario database. 

```{r profile, fig.width=4.5}
aster_profile <- species_profile("Aster lateriflorus", 
                                 ontario_invs,
                                 native = TRUE)
aster_profile

species_profile_plot("Aster lateriflorus", 
                     ontario_invs,
                     native = TRUE)
```

## Data sets

Two tidy data sets of floristic quality data, `chicago` and `missouri`, are included with the ${\tt fqar}$ package. Produced with `assessment_list_glance()`, these show summary information for every floristic quality assessment that used databases 63 and 149, respectively, prior to August 14, 2022. These sets may be useful for visualization or machine-learning purposes. For instance, one might consider the relationship between richness and native mean C in sites assessed using the 2015 Missouri database:

```{r missouri plot, warning = FALSE, fig.width=4.5}
ggplot(missouri, aes(x = native_species, 
                     y = native_mean_c)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(trans = "log10") +
  labs(x = "Native Species (logarithmic scale)",
       y = "Native Mean C") +
  theme_minimal()
```

